---
title: "Multinomial Logit Model"
author: "Sanjit Kangovi"
date: today
---


This assignment explores two methods for estimating the MNL model: (1) via Maximum Likelihood, and (2) via a Bayesian approach using a Metropolis-Hastings MCMC algorithm. 


## 1. Likelihood for the Multi-nomial Logit (MNL) Model

Suppose we have $i=1,\ldots,n$ consumers who each select exactly one product $j$ from a set of $J$ products. The outcome variable is the identity of the product chosen $y_i \in \{1, \ldots, J\}$ or equivalently a vector of $J-1$ zeros and $1$ one, where the $1$ indicates the selected product. For example, if the third product was chosen out of 3 products, then either $y=3$ or $y=(0,0,1)$ depending on how we want to represent it. Suppose also that we have a vector of data on each product $x_j$ (eg, brand, price, etc.). 

We model the consumer's decision as the selection of the product that provides the most utility, and we'll specify the utility function as a linear function of the product characteristics:

$$ U_{ij} = x_j'\beta + \epsilon_{ij} $$

where $\epsilon_{ij}$ is an i.i.d. extreme value error term. 

The choice of the i.i.d. extreme value error term leads to a closed-form expression for the probability that consumer $i$ chooses product $j$:

$$ \mathbb{P}_i(j) = \frac{e^{x_j'\beta}}{\sum_{k=1}^Je^{x_k'\beta}} $$

For example, if there are 3 products, the probability that consumer $i$ chooses product 3 is:

$$ \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{e^{x_1'\beta} + e^{x_2'\beta} + e^{x_3'\beta}} $$

A clever way to write the individual likelihood function for consumer $i$ is the product of the $J$ probabilities, each raised to the power of an indicator variable ($\delta_{ij}$) that indicates the chosen product:

$$ L_i(\beta) = \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} = \mathbb{P}_i(1)^{\delta_{i1}} \times \ldots \times \mathbb{P}_i(J)^{\delta_{iJ}}$$

Notice that if the consumer selected product $j=3$, then $\delta_{i3}=1$ while $\delta_{i1}=\delta_{i2}=0$ and the likelihood is:

$$ L_i(\beta) = \mathbb{P}_i(1)^0 \times \mathbb{P}_i(2)^0 \times \mathbb{P}_i(3)^1 = \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{\sum_{k=1}^3e^{x_k'\beta}} $$

The joint likelihood (across all consumers) is the product of the $n$ individual likelihoods:

$$ L_n(\beta) = \prod_{i=1}^n L_i(\beta) = \prod_{i=1}^n \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} $$

And the joint log-likelihood function is:

$$ \ell_n(\beta) = \sum_{i=1}^n \sum_{j=1}^J \delta_{ij} \log(\mathbb{P}_i(j)) $$



## 2. Simulate Conjoint Data

We will simulate data from a conjoint experiment about video content streaming services. We elect to simulate 100 respondents, each completing 10 choice tasks, where they choose from three alternatives per task. For simplicity, there is not a "no choice" option; each simulated respondent must select one of the 3 alternatives. 

Each alternative is a hypothetical streaming offer consistent of three attributes: (1) brand is either Netflix, Amazon Prime, or Hulu; (2) ads can either be part of the experience, or it can be ad-free, and (3) price per month ranges from \$4 to \$32 in increments of \$4.

The part-worths (ie, preference weights or beta parameters) for the attribute levels will be 1.0 for Netflix, 0.5 for Amazon Prime (with 0 for Hulu as the reference brand); -0.8 for included adverstisements (0 for ad-free); and -0.1*price so that utility to consumer $i$ for hypothethical streaming service $j$ is 

$$
u_{ij} = (1 \times Netflix_j) + (0.5 \times Prime_j) + (-0.8*Ads_j) - 0.1\times Price_j + \varepsilon_{ij}
$$

where the variables are binary indicators and $\varepsilon$ is Type 1 Extreme Value (ie, Gumble) distributed.

The following code provides the simulation of the conjoint data.

:::: {.callout-note collapse="true"}
```{r}
# set seed for reproducibility
set.seed(123)

# define attributes
brand <- c("N", "P", "H") # Netflix, Prime, Hulu
ad <- c("Yes", "No")
price <- seq(8, 32, by=4)

# generate all possible profiles
profiles <- expand.grid(
    brand = brand,
    ad = ad,
    price = price
)
m <- nrow(profiles)

# assign part-worth utilities (true parameters)
b_util <- c(N = 1.0, P = 0.5, H = 0)
a_util <- c(Yes = -0.8, No = 0.0)
p_util <- function(p) -0.1 * p

# number of respondents, choice tasks, and alternatives per task
n_peeps <- 100
n_tasks <- 10
n_alts <- 3

# function to simulate one respondent’s data
sim_one <- function(id) {
  
    datlist <- list()
    
    # loop over choice tasks
    for (t in 1:n_tasks) {
        
        # randomly sample 3 alts (better practice would be to use a design)
        dat <- cbind(resp=id, task=t, profiles[sample(m, size=n_alts), ])
        
        # compute deterministic portion of utility
        dat$v <- b_util[dat$brand] + a_util[dat$ad] + p_util(dat$price) |> round(10)
        
        # add Gumbel noise (Type I extreme value)
        dat$e <- -log(-log(runif(n_alts)))
        dat$u <- dat$v + dat$e
        
        # identify chosen alternative
        dat$choice <- as.integer(dat$u == max(dat$u))
        
        # store task
        datlist[[t]] <- dat
    }
    
    # combine all tasks for one respondent
    do.call(rbind, datlist)
}

# simulate data for all respondents
conjoint_data <- do.call(rbind, lapply(1:n_peeps, sim_one))

# remove values unobservable to the researcher
conjoint_data <- conjoint_data[ , c("resp", "task", "brand", "ad", "price", "choice")]

# clean up
rm(list=setdiff(ls(), "conjoint_data"))
```
::::



## 3. Preparing the Data for Estimation

The "hard part" of the MNL likelihood function is organizing the data, as we need to keep track of 3 dimensions (consumer $i$, covariate $k$, and product $j$) instead of the typical 2 dimensions for cross-sectional regression models (consumer $i$ and covariate $k$). The fact that each task for each respondent has the same number of alternatives (3) helps.  In addition, we need to convert the categorical variables for brand and ads into binary variables.

<!-- _todo: reshape and prep the data_ -->

### Data Preparation

::: {.cell}
```{r}
#| echo: false
#| include: false
#| code-fold: true
#| code-summary: "Load libraries"
library(tidyverse)
# Removed janitor package to avoid missing package error
```
:::

::: {.cell}
```{r}
#| echo: false
#| include: false
#| code-fold: true
#| code-summary: "Read the dataset"
# Read in the conjoint data
df_raw <- read_csv("/Users/sanjitkangovi/Desktop/mysite/blog/project5/conjoint_data.csv")

# View the first few rows
head(df_raw)
```
:::



::: {.cell}
```{r}
#| code-fold: true
#| include: false
#| code-summary: "Clean and recode variables"
# Example of recoding or type conversion

# Convert character variables to factors if needed
df_raw <- df_raw %>%
  mutate(across(where(is.character), as.factor))
```
:::

::: {.cell}
```{r}
# Show summary using kable
library(knitr)
summary_table <- summary(df_raw)
kable(as.data.frame(summary_table), caption = "Summary of Cleaned Data")
```
:::





## 4. Estimation via Maximum Likelihood

<!-- _todo: Code up the log-likelihood function._ -->

::: {.cell}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Prepare data and define log-likelihood function"

# Convert categorical variables to numeric if needed
df_clogit <- df_raw %>%
  mutate(
    brand_num = as.numeric(brand),
    ad_num = as.numeric(ad)
  )

# Log-likelihood function for conditional logit model
log_likelihood_clogit <- function(beta, data) {
  # Compute utility: V_ij = X_ij * beta
  utilities <- beta[1] * data$brand_num +
               beta[2] * data$ad_num +
               beta[3] * data$price

  # Get unique task IDs (choice sets)
  tasks <- unique(data$task)
  log_lik <- 0

  for (t in tasks) {
    subset <- data[data$task == t, ]
    V <- utilities[data$task == t]
    
    # Get utility of chosen alternative
    chosen <- subset$choice == 1
    V_chosen <- V[chosen]
    
    # Calculate log-likelihood contribution
    log_lik <- log_lik + V_chosen - log(sum(exp(V)))
  }

  return(-log_lik)  # Return negative log-likelihood for minimization
}
```
:::

<!-- _todo: Use `optim()` in R or `scipy.optimize()` in Python to find the MLEs for the 4 parameters ($\beta_\text{netflix}$, $\beta_\text{prime}$, $\beta_\text{ads}$, $\beta_\text{price}$), as well as their standard errors (from the Hessian). For each parameter construct a 95% confidence interval._ -->

::: {.cell}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Estimate MLEs using safe vectorized log-likelihood"

library(tidyverse)
library(knitr)
library(MASS)

# Step 1: Ensure choice is numeric
df_raw <- df_raw %>%
  mutate(choice = as.numeric(choice))

# Step 2: Create model matrix (drop intercept)
X <- model.matrix(~ brand + ad + price, data = df_raw)[, -1]

# Step 3: Store target vector and task IDs
y <- df_raw$choice
task <- df_raw$task

# Step 4: Define a valid log-likelihood function that returns a scalar
log_likelihood_mle <- function(beta) {
  utility <- as.vector(X %*% beta)

  df_util <- data.frame(
    task = task,
    utility = utility,
    choice = y
  )

  loglik_vector <- df_util %>%
    group_by(task) %>%
    mutate(
      exp_u = exp(utility),
      denom = sum(exp_u),
      p = exp_u / denom
    ) %>%
    filter(choice == 1) %>%
    pull(p) %>%
    log()

  total_loglik <- sum(loglik_vector)

  return(-total_loglik)  # scalar for optim()
}

# Step 5: Estimate via optim
mle_result <- optim(
  par = rep(0, ncol(X)),
  fn = log_likelihood_mle,
  hessian = TRUE,
  method = "BFGS"
)

# Step 6: Extract estimates and compute SEs
estimates <- mle_result$par
hessian <- mle_result$hessian

se <- if (det(hessian) != 0) {
  sqrt(diag(solve(hessian)))
} else {
  sqrt(diag(ginv(hessian)))
}

# Step 7: Compute 95% CIs
lower <- estimates - 1.96 * se
upper <- estimates + 1.96 * se

# Step 8: Output summary table
mle_summary <- data.frame(
  Parameter = colnames(X),
  Estimate  = round(estimates, 4),
  Std_Error = round(se, 4),
  CI_Lower  = round(lower, 4),
  CI_Upper  = round(upper, 4)
)

kable(mle_summary, caption = "Vectorized MLEs and 95% Confidence Intervals")

```
:::




## 5. Estimation via Bayesian Methods

<!-- _todo: code up a metropolis-hasting MCMC sampler of the posterior distribution. Take 11,000 steps and throw away the first 1,000, retaining the subsequent 10,000._

_hint: Use N(0,5) priors for the betas on the binary variables, and a N(0,1) prior for the price beta._




_hint: instead of calculating post=lik*prior, you can work in the log-space and calculate log-post = log-lik + log-prior (this should enable you to re-use your log-likelihood function from the MLE section just above)_

_hint: King Markov (in the video) use a candidate distribution of a coin flip to decide whether to move left or right among his islands.  Unlike King Markov, we have 4 dimensions (because we have 4 betas) and our dimensions are continuous.  So, use a multivariate normal distribution to pospose the next location for the algorithm to move to. I recommend a MNV(mu, Sigma) where mu=c(0,0,0,0) and sigma has diagonal values c(0.05, 0.05, 0.05, 0.005) and zeros on the off-diagonal.  Since this MVN has no covariances, you can sample each dimension independently (so 4 univariate normals instead of 1 multivariate normal), where the first 3 univariate normals are N(0,0.05) and the last one if N(0,0.005)._ -->

::: {.cell}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Run Metropolis-Hastings MCMC Sampler (11,000 steps)"

library(tidyverse)
library(knitr)
library(MASS)

set.seed(123)  # For reproducibility

# Step 1: Ensure `choice` is numeric
df_raw <- df_raw %>% mutate(choice = as.numeric(choice))

# Step 2: Create design matrix (drop intercept)
X <- model.matrix(~ brand + ad + price, data = df_raw)[, -1]
y <- df_raw$choice
task <- df_raw$task

# Step 3: Log-prior using N(0,5) for binary variables, N(0,1) for price
log_prior <- function(beta) {
  dnorm(beta[1], 0, 5, log = TRUE) +  # beta_netflix
  dnorm(beta[2], 0, 5, log = TRUE) +  # beta_prime
  dnorm(beta[3], 0, 5, log = TRUE) +  # beta_ad
  dnorm(beta[4], 0, 1, log = TRUE)    # beta_price
}

# Step 4: Log-likelihood (same as MLE version, reused)
log_likelihood <- function(beta) {
  utility <- as.vector(X %*% beta)
  df_util <- data.frame(task = task, utility = utility, choice = y)

  loglik_vector <- df_util %>%
    group_by(task) %>%
    mutate(
      exp_u = exp(utility),
      denom = sum(exp_u),
      p = exp_u / denom
    ) %>%
    filter(choice == 1) %>%
    pull(p) %>%
    log()

  return(sum(loglik_vector))  # return total log-likelihood
}

# Step 5: Log-posterior
log_posterior <- function(beta) {
  log_likelihood(beta) + log_prior(beta)
}

# Step 6: Metropolis-Hastings Sampler (11,000 steps, diagonal proposal)
run_mh_sampler <- function(start, n_iter) {
  n_params <- length(start)
  samples <- matrix(NA, nrow = n_iter, ncol = n_params)
  current <- start
  current_logpost <- log_posterior(current)

  # Proposal SDs: N(0, 0.05) for first 3, N(0, 0.005) for price
  proposal_sd <- c(sqrt(0.05), sqrt(0.05), sqrt(0.05), sqrt(0.005))

  for (i in 1:n_iter) {
    proposal <- current + rnorm(n_params, mean = 0, sd = proposal_sd)
    proposal_logpost <- log_posterior(proposal)

    log_accept_ratio <- proposal_logpost - current_logpost

    if (log(runif(1)) < log_accept_ratio) {
      current <- proposal
      current_logpost <- proposal_logpost
    }

    samples[i, ] <- current
  }

  colnames(samples) <- colnames(X)
  return(samples)
}

# Step 7: Run sampler and discard burn-in
mh_samples <- run_mh_sampler(start = rep(0, 4), n_iter = 11000)
mh_posterior <- mh_samples[1001:11000, ]

# Step 8: Summarize posterior
mh_summary <- as.data.frame(mh_posterior) %>%
  summarise(across(everything(), list(
    mean = ~round(mean(.), 4),
    sd = ~round(sd(.), 4),
    lower = ~round(quantile(., 0.025), 4),
    upper = ~round(quantile(., 0.975), 4)
  ), .names = "{.col}_{.fn}"))

kable(t(mh_summary), caption = "Posterior Means, SDs, and 95% Credible Intervals")

```
:::

<!-- _todo: for at least one of the 4 parameters, show the trace plot of the algorithm, as well as the histogram of the posterior distribution._ -->

::: {.cell}
::: {.cell}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Trace plot and histogram for β_price"

library(ggplot2)

# Extract β_price samples
beta_price_draws <- mh_posterior[, 4]
draw_ids <- 1:length(beta_price_draws)

# Trace plot
trace_plot <- ggplot(data.frame(iter = draw_ids, value = beta_price_draws),
                     aes(x = iter, y = value)) +
  geom_line(color = "#6A7B8C", alpha = 0.8) +
  labs(title = "Trace Plot for β_price", x = "Iteration", y = "β_price") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")  # centered, bold
  )

# Histogram
hist_plot <- ggplot(data.frame(value = beta_price_draws), aes(x = value)) +
  geom_histogram(bins = 40, fill = "#6A7B8C", color = "white", alpha = 0.9) +
  labs(title = "Posterior Distribution of β_price", x = "β_price", y = "Frequency") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")  # centered, bold
  )

# Print both plots separately
print(trace_plot)
print(hist_plot)


```
:::

<!-- _todo: report the 4 posterior means, standard deviations, and 95% credible intervals and compare them to your results from the Maximum Likelihood approach._ -->

::: {.cell}
```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Compare MCMC Posterior to MLE Estimates"

# Create tidy MCMC summary
mcmc_summary <- as.data.frame(mh_posterior) %>%
  summarise(across(everything(), list(
    mean  = ~mean(.),
    sd    = ~sd(.),
    lower = ~quantile(., 0.025),
    upper = ~quantile(., 0.975)
  ))) %>%
  pivot_longer(cols = everything(),
               names_to = c("Parameter", ".value"),
               names_sep = "_")

# Assume you already have these from MLE block
mle_summary_df <- data.frame(
  Parameter = colnames(X),
  MLE_Estimate = estimates,
  MLE_SE = se,
  MLE_Lower = estimates - 1.96 * se,
  MLE_Upper = estimates + 1.96 * se
)

# Combine summaries for comparison
comparison_df <- mcmc_summary %>%
  left_join(mle_summary_df, by = "Parameter") %>%
  mutate(across(where(is.numeric), round, 4))

# Display combined table
library(knitr)
kable(comparison_df,
      caption = "Comparison of MCMC Posterior and MLE Estimates",
      col.names = c("Parameter", "Posterior Mean", "Posterior SD", 
                    "95% CI Lower", "95% CI Upper",
                    "MLE Estimate", "MLE SE", 
                    "MLE 95% CI Lower", "MLE 95% CI Upper"))
```
:::

## 6. Discussion

<!-- _todo: Suppose you did not simulate the data. What do you observe about the parameter estimates? What does $\beta_\text{Netflix} > \beta_\text{Prime}$ mean? Does it make sense that $\beta_\text{price}$ is negative?_ -->

::: {.callout-note title="Interpretation of Parameter Estimates"}

Assuming the data reflect real consumer choices (not simulation), the estimated parameters are meaningful:

- **$\beta_\text{Netflix} > \beta_\text{Prime}$**  
  → Consumers prefer **Netflix** over **Prime**, holding all else constant. This means Netflix contributes more to utility than Prime does.

- **$\beta_\text{price} < 0$**  
  → A negative coefficient for **price** is expected: as price increases, the probability of a subscription being chosen decreases — consistent with economic theory and downward-sloping demand.

Together, these suggest the model captures real consumer preferences for brand and sensitivity to price. The direction and relative size of the coefficients align with rational behavior.
:::


<!-- _todo: At a high level, discuss what change you would need to make in order to simulate data from --- and estimate the parameters of --- a multi-level (aka random-parameter or hierarchical) model. This is the model we use to analyze "real world" conjoint data._ -->

::: {.callout-note title="Extending to a Multi-Level (Hierarchical) Logit Model"}

To simulate data from — and estimate parameters of — a **multi-level (random parameter)** model, we would need to:

### 1. Introduce respondent-level heterogeneity
Rather than assuming that all individuals share the same set of coefficients $\boldsymbol{\beta}$, we allow each respondent $i$ to have their own $\boldsymbol{\beta}_i$. These are drawn from a population-level distribution:

$$
\boldsymbol{\beta}_i \sim \mathcal{N}(\boldsymbol{\mu}, \boldsymbol{\Sigma})
$$

Here, $\boldsymbol{\mu}$ is the **average** effect across respondents, and $\boldsymbol{\Sigma}$ captures the **variability** in preferences.

---

### 2. Modify the data generation process
Instead of applying one $\boldsymbol{\beta}$ to all choice tasks, you would:
- Sample a unique $\boldsymbol{\beta}_i$ for each respondent
- Use that $\boldsymbol{\beta}_i$ to generate utilities and choices for their tasks

---

### 3. Use hierarchical Bayesian estimation
To estimate this model, you typically:
- Place priors on both $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$
- Use **Gibbs sampling** or **Hamiltonian Monte Carlo (HMC)** to draw from the full joint posterior
- Software like `Stan`, `PyMC`, or `JAGS` is often used for this

---

### Why it matters:
This hierarchical approach is essential in real-world conjoint analysis because it:
- Accounts for individual differences in preferences
- Provides more accurate predictions at the individual level
- Prevents misleading inferences from population-averaged models

:::












