---
title: "From Distance to Decision: A Study of K-Means and KNN"
author: "Sanjit Kangovi"
date: today
---

<!-- _todo: do two analyses.  Do one of either 1a or 1b, AND one of either 2a or 2b._ -->

## 1a. K-Means

<!-- _todo: write your own code to implement the k-means algorithm.  Make plots of the various steps the algorithm takes so you can "see" the algorithm working.  Test your algorithm on the Palmer Penguins dataset, specifically using the bill length and flipper length variables.  Compare your results to the built-in `kmeans` function in R or Python._ -->
```{r}
#| message: false
#| warning: false
#| echo: true
#| code-fold: true
#| code-summary: "Load Libraries"
library(tidyverse)
library(palmerpenguins)

penguins_data <- read_csv("/Users/sanjitkangovi/Desktop/mysite/blog/project6/palmer_penguins.csv")
  select(bill_length_mm, flipper_length_mm) %>%
  drop_na()

X <- as.matrix(penguins_data)
```

```{r}
#| message: false
#| warning: false
#| echo: true
#| code-fold: true
#| code-summary: "Kmeans functions"
# Helper: Euclidean distance
euclidean <- function(a, b) {
  sqrt(sum((a - b)^2))
}

assign_clusters <- function(X, centroids) {
  apply(X, 1, function(point) {
    dists <- apply(centroids, 1, function(centroid) euclidean(point, centroid))
    which.min(dists)
  })
}

# Recalculate centroids
update_centroids <- function(X, clusters, k) {
  sapply(1:k, function(i) colMeans(X[clusters == i, , drop = FALSE])) %>% t()
}
```
```{r}
#| label: kmeans-animation
#| echo: true
#| code-fold: true
#| code-summary: "K-Means clustering as a GIF"
#| message: false
#| warning: false
#| results: "markup"

# Load libraries
library(tidyverse)
library(gganimate)
library(gifski)
library(palmerpenguins)

# Helper: Euclidean distance
euclidean <- function(a, b) {
  sqrt(sum((a - b)^2))
}

# Assign points to nearest centroid
assign_clusters <- function(X, centroids) {
  apply(X, 1, function(point) {
    dists <- apply(centroids, 1, function(centroid) euclidean(point, centroid))
    which.min(dists)
  })
}

# Recalculate centroids
update_centroids <- function(X, clusters, k) {
  sapply(1:k, function(i) colMeans(X[clusters == i, , drop = FALSE])) %>% t()
}

# Prepare data
X <- na.omit(penguins[, c("bill_length_mm", "flipper_length_mm")])
X <- as.matrix(X)
storage.mode(X) <- "numeric"

# K-means parameters
set.seed(123)
k <- 3
n_iter <- 5
init_idx <- sample(1:nrow(X), k)
centroids <- X[init_idx, ]

# Perform clustering iterations
all_iters <- list()

for (i in 1:n_iter) {
  clusters <- assign_clusters(X, centroids)
  df <- as.data.frame(X)
  colnames(df) <- c("bill_length_mm", "flipper_length_mm")
  df$cluster <- factor(clusters)
  df$iter <- i
  centroids <- update_centroids(X, clusters, k)
  all_iters[[i]] <- df
}

# Combine all iterations
anim_df <- bind_rows(all_iters)

# Create animation plot
p <- ggplot(anim_df, aes(x = bill_length_mm, y = flipper_length_mm, color = cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = "K-Means Iteration: {closest_state}",
       x = "Bill Length (mm)",
       y = "Flipper Length (mm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14)
  ) +
  transition_states(iter, transition_length = 1, state_length = 1) +
  ease_aes('linear')

# Render animation and save as GIF
animate(p, renderer = gifski_renderer("kmeans.gif"), width = 600, height = 500)



```{r}
#| label: builtin-kmeans
#| echo: true
#| code-fold: true
#| code-summary: "Built-in KMeans Plot"
#| message: false
#| warning: false
#| results: "markup"

kmeans_builtin <- kmeans(X, centers = 3, nstart = 10)

df_builtin <- data.frame(X, cluster = factor(kmeans_builtin$cluster))
centroids_builtin <- as.data.frame(kmeans_builtin$centers)
centroids_builtin$cluster <- factor(1:k)

ggplot(df_builtin, aes(x = bill_length_mm, y = flipper_length_mm, color = cluster)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_point(data = centroids_builtin, aes(x = bill_length_mm, y = flipper_length_mm),
             color = "black", size = 4, shape = 8) +
  labs(title = "Built-in KMeans Result",
       x = "Bill Length (mm)", y = "Flipper Length (mm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14)
  )
```

::: {.callout-note title="Interpretation: Built-in KMeans Clustering Result"}

This plot displays the output of R's built-in `kmeans()` function applied to the penguins dataset, using `bill_length_mm` and `flipper_length_mm` as clustering variables with `k = 3`.

- Each point is colored based on its assigned cluster.
- Centroids are shown as black asterisk markers.
- The clusters appear well-separated in 2D space, illustrating how `kmeans()` partitions data based on Euclidean distance.

This plot serves as a benchmark to validate the results of the custom K-Means implementation. The similarity of cluster shapes and centroids between this and the manual version demonstrates that the algorithm was correctly implemented.
:::


<!-- _todo: Calculate both the within-cluster-sum-of-squares and silhouette scores (you can use built-in functions to do so) and plot the results for various numbers of clusters (ie, K=2,3,...,7). What is the "right" number of clusters as suggested by these two metrics?_ -->
```{r}
# Load libraries
library(cluster)
library(ggplot2)
library(gridExtra)

# Compute WCSS and silhouette
wcss <- numeric()
silhouette_avg <- numeric()
k_values <- 2:7

for (k in k_values) {
  km <- kmeans(X, centers = k, nstart = 10)
  wcss[k - 1] <- km$tot.withinss
  sil <- silhouette(km$cluster, dist(X))
  silhouette_avg[k - 1] <- mean(sil[, 3])
}

df_metrics <- data.frame(
  k = k_values,
  WCSS = wcss,
  Silhouette = silhouette_avg
)

# Elbow plot
p1 <- ggplot(df_metrics, aes(x = k, y = WCSS)) +
  geom_line(color = "#5B8FA8", linewidth = 1) +
  geom_point(size = 2, color = "#5B8FA8") +
  labs(title = "Elbow Method", x = "K", y = "WCSS") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "#DADADA")
  )

# Silhouette plot
p2 <- ggplot(df_metrics, aes(x = k, y = Silhouette)) +
  geom_line(color = "#5B8FA8", linewidth = 1) +
  geom_point(size = 2, color = "#5B8FA8") +
  labs(title = "Silhouette Score", x = "K", y = "Avg Silhouette") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "#DADADA")
  )

# Arrange side by side
grid.arrange(p1, p2, ncol = 2)


# Arrange side by side
grid.arrange(p1, p2, ncol = 2)


::: {.callout-note title="Interpretation: Elbow & Silhouette Scores"}

These two side-by-side plots help assess the optimal number of clusters (`K`) for the K-Means algorithm using two complementary metrics:

- **Elbow Method (Left Plot):**
  - This plot shows the Within-Cluster Sum of Squares (WCSS) for values of `K` from 2 to 7.
  - As `K` increases, WCSS decreases, but at a diminishing rate.
  - The "elbow" — where the rate of improvement drops off — suggests an appropriate number of clusters. In this case, the elbow appears to occur around **K = 3 or 4**, indicating diminishing returns beyond that.

- **Silhouette Score (Right Plot):**
  - This plot shows the average silhouette width for each value of `K`, measuring how well-separated the clusters are.
  - Higher scores indicate more distinct and compact clusters.
  - The highest silhouette score may point to **K = 3** as the most natural grouping for this dataset.

Together, both metrics suggest **K = 3** is a strong candidate for the optimal number of clusters.
:::



<!-- _If you want a challenge, add your plots as an animated gif on your website so that the result looks something like [this](https://www.youtube.com/shorts/XCsoWZU9oN8)._ -->







## 2a. K Nearest Neighbors

<!-- _todo: use the following code (or the python equivalent) to generate a synthetic dataset for the k-nearest neighbors algorithm.  The code generates a dataset with two features, `x1` and `x2`, and a binary outcome variable `y` that is determined by whether `x2` is above or below a wiggly boundary defined by a sin function._ -->

```{r}
# gen data -----
set.seed(42)
n <- 100
x1 <- runif(n, -3, 3)
x2 <- runif(n, -3, 3)
x <- cbind(x1, x2)

# define a wiggly boundary
boundary <- sin(4*x1) + x1
y <- ifelse(x2 > boundary, 1, 0) |> as.factor()
dat <- data.frame(x1 = x1, x2 = x2, y = y)
```

```{r}
#| label: gen-knn-data
#| echo: true
#| code-fold: true
#| code-summary: "Generate synthetic data for KNN"
#| message: false
#| warning: false
#| results: "markup"

# Generate data
set.seed(42)
n <- 100
x1 <- runif(n, -3, 3)
x2 <- runif(n, -3, 3)
x <- cbind(x1, x2)

# Define boundary
boundary <- sin(4 * x1) + x1
y <- ifelse(x2 > boundary, 1, 0) |> as.factor()

# Combine into data frame
dat <- data.frame(x1 = x1, x2 = x2, y = y)
```



<!-- _todo: plot the data where the horizontal axis is `x1`, the vertical axis is `x2`, and the points are colored by the value of `y`.  You may optionally draw the wiggly boundary._ -->

```{r}
#| label: plot-knn-synthetic
#| echo: true
#| code-fold: true
#| code-summary: "Plot x1 vs x2 colored by y"
#| message: false
#| warning: false
#| results: "markup"

library(ggplot2)

ggplot(dat, aes(x = x1, y = x2, color = y)) +
  geom_point(size = 2, alpha = 0.8) +
  stat_function(fun = function(x) sin(4 * x) + x,
                color = "black", linetype = "dashed", linewidth = 1) +
  labs(title = "Synthetic Dataset for KNN",
       subtitle = "Dashed line = Decision Boundary (sin(4x1) + x1)",
       x = "x1", y = "x2", color = "y") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5)
  )
```



<!-- _todo: generate a test dataset with 100 points, using the same code as above but with a different seed._ -->

```{r}
#| label: gen-knn-test
#| echo: true
#| code-fold: true
#| code-summary: "Generate test dataset (new seed)"
#| message: false
#| warning: false
#| results: "markup"

# Generate test data with different seed
set.seed(99)
n_test <- 100
x1_test <- runif(n_test, -3, 3)
x2_test <- runif(n_test, -3, 3)
boundary_test <- sin(4 * x1_test) + x1_test
y_test <- ifelse(x2_test > boundary_test, 1, 0) |> as.factor()

# Combine into test data frame
test_data <- data.frame(x1 = x1_test, x2 = x2_test, y = y_test)

# Preview
head(test_data)
```


<!-- _todo: implement KNN by hand.  Check you work with a built-in function -- eg, `class::knn()` or `caret::train(method="knn")` in R, or scikit-learn's `KNeighborsClassifier` in Python._ -->

```{r}
#| label: knn-manual
#| echo: true
#| code-fold: true
#| code-summary: "Manual KNN implementation (k = 5)"
#| message: false
#| warning: false
#| results: "markup"

# Manual KNN function
euclidean_dist <- function(x1, x2) {
  sqrt(sum((x1 - x2)^2))
}

knn_predict <- function(train_X, train_y, test_X, k = 5) {
  preds <- sapply(1:nrow(test_X), function(i) {
    dists <- apply(train_X, 1, function(x) euclidean_dist(x, test_X[i, ]))
    neighbors <- order(dists)[1:k]
    majority <- names(sort(table(train_y[neighbors]), decreasing = TRUE))[1]
    return(majority)
  })
  factor(preds, levels = levels(train_y))
}

# Prepare feature matrices
train_X <- dat[, c("x1", "x2")]
train_y <- dat$y
test_X <- test_data[, c("x1", "x2")]
test_y <- test_data$y

# Predict manually
manual_preds <- knn_predict(train_X, train_y, test_X, k = 5)

# Accuracy
manual_acc <- mean(manual_preds == test_y)
manual_acc
```

```{r}
#| label: knn-built-in
#| echo: true
#| code-fold: true
#| code-summary: "Check with class::knn()"
#| message: false
#| warning: false
#| results: "markup"

library(class)

# Built-in knn prediction
builtin_preds <- knn(train = train_X,
                     test = test_X,
                     cl = train_y,
                     k = 5)

# Accuracy
builtin_acc <- mean(builtin_preds == test_y)
builtin_acc
```


<!-- _todo: run your function for k=1,...,k=30, each time noting the percentage of correctly-classified points from the test dataset. Plot the results, where the horizontal axis is 1-30 and the vertical axis is the percentage of correctly-classified points. What is the optimal value of k as suggested by your plot?_  -->

```{r}
#| label: knn-accuracy-curve
#| echo: true
#| code-fold: true
#| code-summary: "Test accuracy across k = 1 to 30"
#| message: false
#| warning: false
#| results: "markup"

# Accuracy for each k
k_vals <- 1:30
acc_vals <- numeric(length(k_vals))

for (i in seq_along(k_vals)) {
  k <- k_vals[i]
  preds_k <- knn_predict(train_X, train_y, test_X, k)
  acc_vals[i] <- mean(preds_k == test_y)
}

# Create dataframe
knn_perf <- data.frame(k = k_vals, accuracy = acc_vals * 100)

# Plot with subtle color styling
library(ggplot2)

ggplot(knn_perf, aes(x = k, y = accuracy)) +
  geom_line(color = "#5B8FA8", linewidth = 1) +
  geom_point(color = "#5B8FA8", size = 2) +
  labs(title = "KNN Test Accuracy vs. Number of Neighbors",
       x = "Number of Neighbors (k)",
       y = "Test Accuracy (%)") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "#DADADA"),
    panel.grid.minor = element_blank()
  )
```


::: {.callout-note title="Summary: K-Nearest Neighbors (KNN) Classification"}

This section explored the implementation and evaluation of the K-Nearest Neighbors (KNN) algorithm using a synthetically generated dataset with a non-linear decision boundary.

####  Data Generation
A synthetic training dataset with two features (`x1`, `x2`) and a binary outcome `y` was created. The outcome was determined by whether `x2` was above a wiggly boundary defined by the function `sin(4x1) + x1`. A separate test dataset was generated using the same logic but a different random seed.

####  Manual KNN Implementation
We implemented the KNN algorithm from scratch by:
- Calculating Euclidean distances between test points and all training points
- Identifying the `k` nearest neighbors
- Assigning class labels via majority vote

The manual implementation was validated by comparing its accuracy to the built-in `class::knn()` function from R. The results matched exactly, confirming correctness.

####  Model Evaluation
The manual KNN function was run for `k = 1` to `30`, and the classification accuracy on the test set was recorded. These results were plotted to visualize how accuracy varies with `k`. The plot revealed that performance peaks at a particular value of `k`, suggesting an optimal number of neighbors for this dataset.

####  Interpretation
- Low values of `k` tend to overfit, capturing noise in the training data.
- High values of `k` can underfit, smoothing over important decision boundaries.
- The optimal `k` balances these two effects and yielded the highest accuracy in our test.

This section demonstrates a full machine learning pipeline: synthetic data generation, custom model implementation, model validation, and hyperparameter tuning through visualization.
:::









